Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-11 response-to-reviews.Rmd

**Done**

line165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this. Please mention along with at least 1 reference.

line185: Buffering again. I supposed what is 'buffering' for the authors would conventionally be thought of differently by meteorologists. Please consult with one of them, and consider using different wording.

*Buffering was replaced with the following sentences:*
*Wind speeds are also higher at the top of the canopy, that are substantially reduced moving into lower canopy layers due to resistance caused by canopy foliage  [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].*

@@

**Done**

Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.

Figure 3: Humid and drought is not a straight comparison. Well-watered vs. drought would be clearer as there are some assumptions about gs in these conditions that underlie the modelling being done here.

*This is either a semantics issue, or a case where the reviewer is expecting greater realism that what we're actually doing. You should make that clear in the caption (also addresses comment above), and perhaps use "humid" and "dry" (rather than drought).

Also note that well-watered would refer to soil moisture, and require modeling water transport through roots and stem. Drought could refer to either soil dryness or atmospheric dryness. In my mind, it's always been the latter.*

*changed to "drought" to "dry"

### 2022-01-11 response-to-reviews.Rmd

In the first part of Section 5.1, the authors conclude this paragraph stating the 'dominant role of vertical profiles in microclimate in shaping tree growth rates'. However, I believe the text has confused the very large effect of light with other aspects of forest microclimate, and they very much need to clarify this. If they believe other aspects of microclimate apart from light are drivers, then I believe this would be a minority opinion in the forest biology community. If they wish to speculate about other microclimate factors that cannot be disentangled from light, I suggest that they clearly indicate their opinion and speculation. It is perhaps an area for future research to disentangle light microclimate from aspects of vertical microclimate.

*[issue #91](https://github.com/EcoClimLab/vertical-thermal-review/issues/91)*

I am uncomfortable reading some of the major conclusions of the work here, such as speculation that large canopy trees are most vulnerable to warming when water is limited, but understory trees may be more vulnerable [to warming] when well-watered. This is not consistent with my knowledge of 'preponderance of available data'. Perhaps the authors implying that understory trees will succumb to warming earlier in a heat wave than large canopy trees? Is this based on the T50 evidence (l. 542-548) which is from 2 studies, or from one tree-ring study (l. 674-676), or both? An intriguing thought, but these studies and the corresponding author’s study have weaknesses and no clean experiment on this has been set up and done. I recommend that the more cautious language on l. 676-678 should in fact be repeated in the conclusions relevant to this point. It’s less stimulating and sober, but true, that it’s difficult to say so with strength of evidence and remains for further testing with good experimental design. An objective of such a review can stimulate further and better research.

**We agree that more caution is warrented on comparing canopy vs understory responses to warming, particularly because understory responses are highly uncertain. We have reworded relevant text as follows:**

- **"Implications: Global Change Responses : Warming" section, 4th par (previous lines 793-797: "While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), *limited tree-ring evidence* indicates that understory trees *can* exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig 4b, @rollinsonClimateSensitivityUnderstory2020]."**

### 2022-01-11 vertical-thermal-review.Rmd
starting on issue #90

Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018].
However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 
Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), tree-ring evidence indicates that understory trees tend to exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the more buffered understory may be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].

>>>>

Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018].
However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 
Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), limited tree-ring evidence indicates that understory trees can exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the more buffered understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
